OldCrystalPlan: an Experiment Planning Tool for Crystallography

### 2017-05-23 docs/CrystalPlan_paper/paper.tex
CrystalPlan: Main window a little wider.
Refs #old change CrystalPlan to OldCrystalPlan for rmp

\title{CrystalPlan: an Experiment Planning Tool for Crystallography}

\shorttitle{CrystalPlan: an Experiment Planning Tool}

@@

\cauthor[]{Janik}{Zikovsky}{zikovskyjl@ornl.gov}{address if different from
\aff}
\author[]{Peter}{Peterson}
\author[]{Xiaoping}{Wang}
\author[]{Matthew}{Frost}
\author[]{Christina}{Hoffmann}

@@

\begin{synopsis}
Describes CrystalPlan, a software program for the automatic creation of 
crystallography experiment plans for time-of-flight diffractometers.
\end{synopsis}

\begin{abstract}
Beam time at large user program based x-ray and neutron scattering facilities is
in high demand and always at a premium. CrystalPlan, a highly efficient
experiment planning software has been developed to maximize the use of available
beamtime per sample per experiment. 
This program can calculate and optimize the data coverage of a crystal in
reciprocal space in a single-crystal diffraction time-of-flight experiment.
CrystalPlan can help a user build an experiment plan that will acquire the most
unique data possible, with sufficient coverage but limited redundancy, therefore
increasing scientific productivity. 
A user friendly GUI (Graphical User Interface) including a 3D viewer, an
automated coverage optimizer, and an option to reorient the crystal for the measurement 
of selected $hkl$s on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
tailored to the needs of the scientist (enough reflections measured for a structure
analysis), but without spending time collecting redundant data, CrystalPlan will
greatly increase the scientific productivity and throughput of TOPAZ and other
similar beamlines. Users will be able to measure samples highly efficiently,
allowing them to explore parameters such as temperature and pressure more fully.

@@

CrystalPlan is designed with these capabilities:

@@

CrystalPlan currently supports area detectors with flat faces and arbitrary
rectangular dimensions; this reflects the detector types in use at SNS (Anger
cameras and 8-pack $^3$He tube detectors), but the software could easily be
extended for more complicated detector shapes as needed. The detector configuration is loaded from
a text file containing the location, size, orientation and number of pixels of
each detector. Other settings to be adjusted for an instrument include the
detectable range of neutron/x-ray wavelength.

@@

CrystalPlan can accomodate a variety of sample
orientation goniometers. A base Goniometer class (in the object-oriented
programming sense) is subclassed for specific types of goniometers, allowing a
unified interface. This base class acts as a \emph{universal goniometer},
which uses the conventional Eulerian goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from the base Goniometer can convert
the desired $\phi$, $\chi$, $\omega$ angles to the internal motor positions it requires. 
The class hierarchy allows for goniometer limitations: for example, TOPAZ is
currently equipped with a goniometer where $\chi$ is fixed at 45 degrees
(at $\omega$ to the incident beam). CrystalPlan handles this by limiting the
sample orientation angles accessible to the user to $\phi$ and $\omega$.

The sample unit cell and orientation is typically determined by the data
collection/reduction software. In our case, from an initial data set, the data
analysis program ISAW (Integrated Scattering Analysis
Workbench \cite{Mikkelson05}) provides us with the unit cell of the sample (via
the B matrix) and the sample mounting orientation (U matrix).
The combined UB matrix is then saved to a
text file along with the lattice parameters. 
CrystalPlan can load this UB matrix file and uses it for reciprocal coverage
calculations and optimization.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. CrystalPlan then
generates a list of $hkl$ reflections with d-spacings larger than the $d_{min}$
specified. For each sample orientation being simulated and for each reflection, a function calculates at
what wavelength $\lambda$ and in which direction it scatters. If $\lambda$ 
is within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

@@

In volume coverage mode, a 4D array is generated where the first 3 dimensions
correspond to {\bf Q}-space, and the $4^{th}$ dimension is a list of $n$
entries. Each entry in the $4^{th}$ dimension points to a voxel that is an
equivalent due to crystal symmetry. This allows CrystalPlan to quickly compute a volume coverage
map that takes into account crystal symmetry and the resulting redundancy: at
each voxel, all $n$ equivalent voxels are checked to see if any of them were
measured.

@@

CrystalPlan allows the user to manually enter lists of sample orientations to
simulate and graphically display the coverage in reciprocal space;
however, finding an optimal plan by trial and error is in general 
time consuming and only feasible for an experienced crytallographer.
To make the process more efficient, CrystalPlan includes an Automatic
Coverage Optimizer, which takes as its inputs the desired number of sample
orientations and uses a genetic algorithm to find a solution that satisfies a
user-specified coverage criterion (for example, the user may ask for $85\%$ of 
Bragg reflections to be measured, or for $90\%$ of {\bf Q}-volume).

@@

To this end, CrystalPlan was written in Python \cite{python}, a powerful, open,
cross-platform scripting language. Python was chosen for its ease of cross-platform deployment, and for
its attractive language features. The numpy and scipy libraries (two
open-source, commonly installed Python libraries) were used for many of the
calculations \cite{numpy,scipy}. The most time-critical calculations
were also written with inline C code, which increases the computation speed by
up to $400\times$ as compared with the pure Python version (which can still be
used, in case of incompatibilities).

@@

The calculation code of CrystalPlan is placed in a model layer and kept
completely separate from eventual GUI elements, using a Model/View-Controller
architecture \cite{Gamma95}. This makes it possible to run calculations using
Python scripts, if desired. Unit tests for each part of the calculations are used to ensure
consistent results on different platforms and  systems.

@@

An attractive and easy-to-use GUI was considered an important part of
CrystalPlan, since the program is meant to be used interactively as a guide to
the experimenter whether new or experienced. This requires intuitive GUI
interfaces and workflows.
The user interface was written using wxPython
\cite{wxPython}, a mature cross-platform GUI toolkit, as well as the Enthought
Traits GUI. Some of the 2D plots use matplotlib \cite{matplotlib},  and the 3D
visualization elements use the Mayavi2 toolkit for Python \cite{mayavi}, itself
based on VTK \cite{VTK}. All of these libraries are  open-source.

@@

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the reflections,
taking into account the crystal Laue symmetry (orthorhombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The eight orientations were run
unattended overnight.

Reflections were integrated using ISAW, and the resulting file was
loaded into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated reflections
found. Good agreement was found; the root-mean-square difference between the
predicted and actual measurement wavelength was only 0.016 \ang. The offsets in the reflection
position were plotted, which helped show that the main contributor to this error
was a slight misalignment of the goniometer.

@@

CrystalPlan was also successfully used in a similar way during experiments run
on the TOPAZ instrument by Yuri Janssen from State University of New York, Stony
Brook, and Henrik Clausen from University of Aarhus, Denmark.

@@

We have described CrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help increase scientific productivity at SNS
beamlines such as TOPAZ. With some relatively minor modifications, the program
could also be adapted for use at reactor sources. Monochromatic diffractometers
could be handled by adding support sweeps consisting of many motor positions;
Laue diffractometers on a constant intensity beam would requiring
detecting and removing reflections with chromatic overlap.

One planned future development for CrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
CrystalPlan would interface with data analysis software such as ISAW and use
recently developed capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular reflections. When the signal intensity
becomes sufficient, the sample will automatically be moved to the next sample
orientation, further adjusting data acquisition time for each specific sample.
Combined with advanced sample changers coming on-line at TOPAZ, this will allow
for extremely efficient, unattended and automated data acquisition.

>>>>

\title{OldCrystalPlan: an Experiment Planning Tool for Crystallography}

\shorttitle{OldCrystalPlan: an Experiment Planning Tool}

@@

\cauthor[]{Janik}{Zikovsky}{zikovskyjl@ornl.gov}{address if different from
\aff}
\author[]{Peter F.}{Peterson}
\author[]{Xiaoping P.}{Wang}
\author[]{Matthew}{Frost}
\author[]{Christina}{Hoffmann}

@@

\begin{synopsis}
Describes OldCrystalPlan, a software program for the automatic creation of 
crystallography experiment plans for time-of-flight diffractometers.
\end{synopsis}

\begin{abstract}
Beam time at large user program based x-ray and neutron scattering facilities is
in high demand and always at a premium. OldCrystalPlan, a highly efficient
experiment planning software has been developed to maximize the use of available
beamtime per sample per experiment. 
This program can calculate and optimize the data coverage of a crystal in
reciprocal space in a single-crystal diffraction time-of-flight experiment.
OldCrystalPlan can help a user build an experiment plan that will acquire the most
unique data possible, with sufficient coverage but limited redundancy, therefore
increasing scientific productivity. 
A user friendly GUI (Graphical User Interface) including a 3D viewer, an
automated coverage optimizer, and an option to reorient the crystal for the measurement 
of selected $hkl$s on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The goal of OldCrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
tailored to the needs of the scientist (enough reflections measured for a structure
analysis), but without spending time collecting redundant data, OldCrystalPlan will
greatly increase the scientific productivity and throughput of TOPAZ and other
similar beamlines. Users will be able to measure samples highly efficiently,
allowing them to explore parameters such as temperature and pressure more fully.

@@

OldCrystalPlan is designed with these capabilities:

@@

OldCrystalPlan currently supports area detectors with flat faces and arbitrary
rectangular dimensions; this reflects the detector types in use at SNS (Anger
cameras and 8-pack $^3$He tube detectors), but the software could easily be
extended for more complicated detector shapes as needed. The detector configuration is loaded from
a text file containing the location, size, orientation and number of pixels of
each detector. Other settings to be adjusted for an instrument include the
detectable range of neutron/x-ray wavelength.

@@

OldCrystalPlan can accomodate a variety of sample
orientation goniometers. A base Goniometer class (in the object-oriented
programming sense) is subclassed for specific types of goniometers, allowing a
unified interface. This base class acts as a \emph{universal goniometer},
which uses the conventional Eulerian goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from the base Goniometer can convert
the desired $\phi$, $\chi$, $\omega$ angles to the internal motor positions it requires. 
The class hierarchy allows for goniometer limitations: for example, TOPAZ is
currently equipped with a goniometer where $\chi$ is fixed at 45 degrees
(at $\omega$ to the incident beam). OldCrystalPlan handles this by limiting the
sample orientation angles accessible to the user to $\phi$ and $\omega$.

The sample unit cell and orientation is typically determined by the data
collection/reduction software. In our case, from an initial data set, the data
analysis program ISAW (Integrated Scattering Analysis
Workbench \cite{Mikkelson05}) provides us with the unit cell of the sample (via
the B matrix) and the sample mounting orientation (U matrix).
The combined UB matrix is then saved to a
text file along with the lattice parameters. 
OldCrystalPlan can load this UB matrix file and uses it for reciprocal coverage
calculations and optimization.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. OldCrystalPlan then
generates a list of $hkl$ reflections with d-spacings larger than the $d_{min}$
specified. For each sample orientation being simulated and for each reflection, a function calculates at
what wavelength $\lambda$ and in which direction it scatters. If $\lambda$ 
is within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

@@

In volume coverage mode, a 4D array is generated where the first 3 dimensions
correspond to {\bf Q}-space, and the $4^{th}$ dimension is a list of $n$
entries. Each entry in the $4^{th}$ dimension points to a voxel that is an
equivalent due to crystal symmetry. This allows OldCrystalPlan to quickly compute a volume coverage
map that takes into account crystal symmetry and the resulting redundancy: at
each voxel, all $n$ equivalent voxels are checked to see if any of them were
measured.

@@

OldCrystalPlan allows the user to manually enter lists of sample orientations to
simulate and graphically display the coverage in reciprocal space;
however, finding an optimal plan by trial and error is in general 
time consuming and only feasible for an experienced crytallographer.
To make the process more efficient, OldCrystalPlan includes an Automatic
Coverage Optimizer, which takes as its inputs the desired number of sample
orientations and uses a genetic algorithm to find a solution that satisfies a
user-specified coverage criterion (for example, the user may ask for $85\%$ of 
Bragg reflections to be measured, or for $90\%$ of {\bf Q}-volume).

@@

To this end, OldCrystalPlan was written in Python \cite{python}, a powerful, open,
cross-platform scripting language. Python was chosen for its ease of cross-platform deployment, and for
its attractive language features. The numpy and scipy libraries (two
open-source, commonly installed Python libraries) were used for many of the
calculations \cite{numpy,scipy}. The most time-critical calculations
were also written with inline C code, which increases the computation speed by
up to $400\times$ as compared with the pure Python version (which can still be
used, in case of incompatibilities).

@@

The calculation code of OldCrystalPlan is placed in a model layer and kept
completely separate from eventual GUI elements, using a Model/View-Controller
architecture \cite{Gamma95}. This makes it possible to run calculations using
Python scripts, if desired. Unit tests for each part of the calculations are used to ensure
consistent results on different platforms and  systems.

@@

An attractive and easy-to-use GUI was considered an important part of
OldCrystalPlan, since the program is meant to be used interactively as a guide to
the experimenter whether new or experienced. This requires intuitive GUI
interfaces and workflows.
The user interface was written using wxPython
\cite{wxPython}, a mature cross-platform GUI toolkit, as well as the Enthought
Traits GUI. Some of the 2D plots use matplotlib \cite{matplotlib},  and the 3D
visualization elements use the Mayavi2 toolkit for Python \cite{mayavi}, itself
based on VTK \cite{VTK}. All of these libraries are  open-source.

@@

The resulting UB matrix was loaded into OldCrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the reflections,
taking into account the crystal Laue symmetry (orthorhombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The eight orientations were run
unattended overnight.

Reflections were integrated using ISAW, and the resulting file was
loaded into OldCrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated reflections
found. Good agreement was found; the root-mean-square difference between the
predicted and actual measurement wavelength was only 0.016 \ang. The offsets in the reflection
position were plotted, which helped show that the main contributor to this error
was a slight misalignment of the goniometer.

@@

OldCrystalPlan was also successfully used in a similar way during experiments run
on the TOPAZ instrument by Yuri Janssen from State University of New York, Stony
Brook, and Henrik Clausen from University of Aarhus, Denmark.

@@

We have described OldCrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help increase scientific productivity at SNS
beamlines such as TOPAZ. With some relatively minor modifications, the program
could also be adapted for use at reactor sources. Monochromatic diffractometers
could be handled by adding support sweeps consisting of many motor positions;
Laue diffractometers on a constant intensity beam would requiring
detecting and removing reflections with chromatic overlap.

One planned future development for OldCrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
OldCrystalPlan would interface with data analysis software such as ISAW and use
recently developed capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular reflections. When the signal intensity
becomes sufficient, the sample will automatically be moved to the next sample
orientation, further adjusting data acquisition time for each specific sample.
Combined with advanced sample changers coming on-line at TOPAZ, this will allow
for extremely efficient, unattended and automated data acquisition.
